---
title: "PISCO_kelp-urchin-sheephead_data"
author: "Jess Hopf"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
library(tidyverse); library(ggpubr)
```

# Background

## Data source
* PISCO Kelp Forest Community Surveys
* Downloaded from [DataONE Data catalog](https://search.dataone.org/view/doi%3A10.6085%2FAA%2FPISCO_kelpforest.1.6) 04/02/2040.
* data will need to be downloaded from catalog before use, or contact author


## Aims of this script
* Obtain measure of mean and variability of kelp, urchin, and sheephead densities in Southern California
* Focus on Channel Islands (San Miguel, Santa Rosa, Anacapa and Santa Cruz) in order to constrain variability in some of the other drivers of kelp abundance including productivity, upwelling, and wave exposure.


# Code 
## Setup and read-in

```{r}
# clear all current variable assignments
rm(list = ls())


# Read in taxon table
KelpForest_Taxon <- read_csv("PISCO_kelpforest_taxon_table.1.2.csv")

# read in sites table
KelpForest_Sites <- read_csv("PISCO_kelpforest_site_table.1.2.csv")

# read in Swath data
KelpForest_Swath <- read_csv("PISCO_kelpforest_swath.1.2.csv")

# read in Quad data
KelpForest_Quad <- read_csv("PISCO_kelpforest_quad.1.2.csv")

# read in fish data
KelpForest_Fish <- read_csv("PISCO_kelpforest_fish.1.3.csv")

```


## Explore data
Add MPA status to data
```{r}
MPA_desg <- KelpForest_Sites %>% select(site, site_status) %>% unique() 

KelpForest_Swath <- left_join(KelpForest_Swath, MPA_desg, by = "site")

KelpForest_Quad <- left_join(KelpForest_Quad, MPA_desg, by = "site")

```



### Kelp adults

Pull out data
```{r}
# filter for  Giant kelp (Macrocystis pyrifera = MACPYRAD)
KFSwatch_Kelp <- KelpForest_Swath %>% filter(classcode == "MACPYRAD")

# deal with NAs for size (number of individual stipes growing for each individual) and count
# no other info given, so will assume = 1
KFSwatch_Kelp <- KFSwatch_Kelp %>% replace_na(list(size = 1, count = 1))

# cut out non-essential columns
KFSwatch_Kelp <- KFSwatch_Kelp %>% select(year:size,site_status) 

# Get stipe density 
KFSwatch_Kelp <- KFSwatch_Kelp %>%
  mutate(Kelp_plants = count,
         Kelp_stipes = count*size) 

# sum within transect
KFSwatch_Kelp <- KFSwatch_Kelp %>%
  group_by(year, site, zone, transect, site_status) %>% 
  summarise(Kelp_plants = sum(Kelp_plants),
         Kelp_stipes = sum(Kelp_stipes))
  

```


Convert stipe density to wet biomass:
-	stipe density to dry biomass (kg) density: y = 0.33x + 0.16 (Reed et al. 2009)
-	wet-to-dry ratio = 0.094 (Rassweiler et al. 2018)

```{r}
KFSwatch_Kelp <- KFSwatch_Kelp %>% 
  mutate(Kelp_biomassdry = 0.33*Kelp_stipes + 0.16,
         Kelp_biomasswet = Kelp_biomassdry/0.094)

```


Focusing on Channel Islands

```{r}
# list of sites
KelpForest_Sites %>% filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% pull(site) %>% unique()

# pull out islands and separate out side
Kelp_df <- KFSwatch_Kelp %>% rowwise() %>% 
  filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% 
  separate_wider_regex(site,c(site = ".*", "_", side = ".*"))


```



#### Scaling transition
- Using Channel Islands
- calculate variance among transects, take mean of years

```{r}
Kelp_var_yr <- Kelp_df %>%
  group_by(year, site, side, zone) %>%
  summarise(Kelp_biomasswet_var = var(Kelp_biomasswet))
  
  # using this one (mean of the mean variance by year)  
  Kelp_var_yr %>% 
  group_by(year) %>%
  summarise(Kelp_biomasswet_meanvar = mean(Kelp_biomasswet_var, na.rm = T)) %>% 
  pull(Kelp_biomasswet_meanvar) %>% mean()
  
  # compare to overall var
  Kelp_var_yr %>% 
  ungroup() %>%
  pull(Kelp_biomasswet_var) %>% mean(., na.rm = T)


```


### Kelp recruits 

```{r}
# filter for Giant kelp recruits
KFQuad_Kelp <- KelpForest_Quad %>% filter(classcode == "MACPYRJUV")


# cut out non-essential columns
KFQuad_Kelp <- KFQuad_Kelp %>% select(year:count,site_status) 


# avg per transect (because multiple transects were taken at different depths)
KFQuad_Kelp <- KFQuad_Kelp %>%
  group_by(year, site, zone, transect, site_status) %>% 
  summarise(Kelp_juvs = mean(count))


# pull out islands and separate out side
KelpR_df <- KFQuad_Kelp %>% rowwise() %>% 
  filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% 
  separate_wider_regex(site,c(site = ".*", "_", side = ".*"))


# Calc year-to-year mean/stdev across all sites
# note data runs 2000 2010
KelpR_dfyr <- KelpR_df %>% 
  group_by(year) %>% 
  summarise(across(Kelp_juvs, list(mean = mean, sd = sd)))


# plot
ggplot(KelpR_df %>% group_by(site, year, site_status) %>% 
      summarise(across(Kelp_juvs, list(mean = mean)))) +
  geom_line(aes(y = Kelp_juvs_mean, x = year, color = site )) +
  theme(legend.position="none") +
  facet_wrap(~site_status, ncol = 1)

ggplot(KelpR_df %>% group_by(site, year) %>% 
      summarise(across(Kelp_juvs, list(mean = mean)))) +
  geom_line(aes(y = Kelp_juvs_mean, x = year, color = site )) +
  theme(legend.position="none") +
  geom_line(data = KelpR_dfyr,
            aes(y = Kelp_juvs_mean, x = year), size = 1) +
  theme(legend.position="none")


#calc yr2yr var ----

# detrend:
    # fit linear model
    Model <- glm(KelpR_dfyr$Kelp_juvs_mean ~ KelpR_dfyr$year)
    
    # mean plus residuals
    tsDT <- mean(KelpR_dfyr$Kelp_juvs_mean) + Model$residuals
    
    # make dataframe
    Model_df <- tibble(year = KelpR_dfyr$year, tsDT)
    
    ggplot(Model_df) + 
      geom_line(aes(y = tsDT, x = year))

# normalise and get varaince
    tsDTnorm <- tsDT/mean(tsDT)   
    
    sd(tsDTnorm)
    

```


### Urchins (aduts & recruits)

#### Urchin adults

Pull out data
```{r}
# filter for adult (>2.5cm test diameter) purple and red (MESFRAAD) urchins 
KFSwatch_Urchins <- KelpForest_Swath %>% filter(classcode %in% c("MESFRAAD", "STRPURAD"))

# summarize size (test diameter, cm; by year, species, across everywhere)
Urchin_Size <- KFSwatch_Urchins %>% 
  filter(!is.na(size)) %>% 
  uncount(weights = count) 

Urchin_Size %>% 
  group_by(year, classcode) %>% 
  summarise(mean(size), median(size), max(size), min(size), n())

Urchin_Size %>% 
  group_by(classcode) %>% 
  summarise(mean(size), median(size), max(size), min(size), n())

# calculate biomass from size (following Ling et al 2015 supp table 1)
# this is the avg kg of an urchin
red_bio <- (0.0499 * 7 + 0.0019)
pur_bio <- (0.0499 * 4.1 + 0.0019)


# cut out non-essential columns
KFSwatch_Urchins <- KFSwatch_Urchins %>% select(year:count, site_status) 

# pull out Islands
KFSwatch_Urchins <- KFSwatch_Urchins %>%
  filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% 
  separate_wider_regex(site,c(site = ".*", "_", side = ".*"))

# Sum within transects
KFSwatch_Urchins <- KFSwatch_Urchins %>%
  group_by(year, site, side, zone, classcode, transect, site_status) %>%
  summarise(count = sum(count))

# pull out spp and rename for joining
Urchins_df <- KFSwatch_Urchins %>% 
  pivot_wider(names_from = classcode, values_from = count) %>% 
  rename(red_count = MESFRAAD, purple_count = STRPURAD) %>% 
  mutate(red_biomass = red_count*red_bio, purple_biomass = purple_count*pur_bio)

# add urchins to get total
Urchins_df <- Urchins_df %>% 
  mutate(Urchins_count = sum(red_count, purple_count, na.rm=T),
         Urchins_biomass = sum(red_biomass, purple_biomass, na.rm=T))



```

#### Scaling transition
- Using Channel Islands
- calculate variance among transects, take mean of years

```{r}
# sep species out again and the summarise ('Urchins' spp = both)
Urchins_var_yr <- Urchins_df %>%
  pivot_longer(cols = c(red_count:Urchins_biomass), names_to = c("spp", ".value"), names_sep = "\\_") %>% 
  group_by(year, site, side, zone, spp) %>%
  summarise(Urchin_biomass_var = var(biomass))
  
  # using this one (mean of the mean variance by year)  
  Urchins_var_yr %>% 
  group_by(year, spp) %>%
  summarise(Urchin_biomass_meanvar = mean(Urchin_biomass_var, na.rm = T)) %>% 
  group_by(spp) %>%  
  summarise(Urchin_biomass_yearvar = mean(Urchin_biomass_meanvar, na.rm = T))
  
  # compare to overall var
  Urchins_var_yr %>% 
  group_by(spp) %>%
  summarise(Urchin_biomass_meanvar = mean(Urchin_biomass_var, na.rm = T))


```




#### Urchin recruits

No urchin recruits collected for the Channel Islands - all are northern and central sites.
We will use this to give us an idea of the magnitude of variance in recruits arriving.

pull out data
```{r}
# filter for recruits (<2.5cm test diameter) purple and red (MESFRAAD) urchins 
# note this only applies for 2014 onwards
KFSwatch_UrchinREC <- KelpForest_Swath %>% filter(classcode %in% c("MESFRAREC", "STRPURREC"))

# cut out non-essential columns
KFSwatch_UrchinREC <- KFSwatch_UrchinREC %>% select(year:count, site_status) 

# Sum within transects
KFSwatch_UrchinREC <- KFSwatch_UrchinREC %>%
  group_by(year, site, zone, classcode, transect, site_status) %>%
  summarise(count = sum(count))


# Calc year-to-year mean/stdev across all transects
KFSwatch_UrchinRECyr <- KFSwatch_UrchinREC %>% 
  group_by(year) %>% 
  summarise(across(count, list(mean = mean, sd = sd)))

# remove heatwave years 2014-16
KFSwatch_UrchinRECyr <- KFSwatch_UrchinRECyr %>% 
  filter(!year %in% c(2014, 2015, 2016))


# plot
ggplot(KFSwatch_UrchinREC %>% group_by(site, year) %>% 
      summarise(across(count, list(mean = mean)))) +
  geom_line(aes(y = count_mean, x = year, color = site )) +
  theme(legend.position="none") +
  geom_line(data = KFSwatch_UrchinRECyr,
            aes(y = count_mean, x = year), size = 1) +
  theme(legend.position="none")


#calc yr2yr var ----
KFSwatch_UrchinRECyr %>% 
  ungroup %>% 
  summarise(across(count_mean:count_sd, list(mean = mean, sd = sd)))


# detrend:
    # fit linear model
    Model <- glm(KFSwatch_UrchinRECyr$count_mean ~ KFSwatch_UrchinRECyr$year)
    
    # mean plus residuals
    tsDT <- mean(KFSwatch_UrchinRECyr$count_mean) + Model$residuals
    
    # make dataframe
    Model_df <- tibble(year = KFSwatch_UrchinRECyr$year, tsDT)
    
    ggplot(Model_df) + 
      geom_line(aes(y = tsDT, x = year))

# normalise and get varaince
    tsDTnorm <- tsDT/mean(tsDT)   
    
    sd(tsDTnorm)

```





### Urchin-Kelp

Join tables
```{r}
KelpUrchin_df <- Kelp_df %>% 
  full_join(Urchins_df,
            by = c("year","site", "side", "zone", "transect", "site_status")) 

# replace NA with 0
KelpUrchin_df <- KelpUrchin_df %>% replace(is.na(.), 0)

# set site status 
KelpUrchin_df <- KelpUrchin_df %>% 
  mutate(site_status = as.factor(site_status))

# remove sites with <3 years of data
  # find sites with >3 years
  site_keep <- KelpUrchin_df %>% 
    select(site, year) %>% 
    unique() %>% 
    group_by(site) %>% 
    filter(n()>2) %>% pull(site)
  
  KelpUrchin_df <- KelpUrchin_df %>% 
    filter(site %in% site_keep)

  
# calc stats at site level
KelpUrchin_df_site <- KelpUrchin_df %>% 
  group_by(year, site, site_status) %>% 
  summarise(across(Kelp_plants:Urchins_biomass, ~mean(., na.rm = T)))

# take log
KelpUrchin_df_site <- KelpUrchin_df_site %>% 
  mutate(across(Kelp_plants:Urchins_biomass, ~log(.), .names = "{col}_ln"))



```


Transect scale

urchins v. kelp
```{r}
# basic
ggplot(KelpUrchin_df) +
  geom_point(aes(x = Urchins_biomass, y = Kelp_biomasswet), alpha = 0.3) +
  ylab('Kelp (kg/60m2)') + xlab('Urchins (kg/60m2)')


# facet MPA status
ggplot(KelpUrchin_df, aes(x = Urchins_biomass, y = Kelp_biomasswet, color = site_status)) +
  geom_point(alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1, scales = 'free') + 
  geom_rug(linewidth=0.1) +
  ylab('Kelp (kg/60m2)') + xlab('Urchins (kg/60m2)')

ggscatterhist(
  KelpUrchin_df, x = "Urchins_biomass", y = "Kelp_biomasswet",
  color = "site_status", fill = "site_status",
  size = 3, alpha = 0.6,
  margin.plot = "boxplot",
  ggtheme = theme_bw())



# facet site (+MPA status)
ggplot(KelpUrchin_df, aes(x = Urchins_biomass, y = Kelp_biomasswet, color = site_status)) +
  geom_point(alpha = 0.3) +
  facet_wrap(vars(site_status, site)) + 
  geom_rug(linewidth=0.1) +
  ylab('Kelp (kg/60m2)') + xlab('Urchins (kg/60m2)')



# facet MPA status + colour urchin
ggplot(KelpUrchin_df) +
  geom_point(aes(x = purple_biomass, y = Kelp_biomasswet), colour = "purple", alpha = 0.3) +
  geom_point(aes(x = red_biomass, y = Kelp_biomasswet), colour = "red", alpha = 0.3) +
    facet_wrap(vars(site_status), ncol = 1, scales = 'free') +
  ylab('Kelp (kg/60m2)') + xlab('Urchins (kg/60m2)')


ggplot(KelpUrchin_df) +
  geom_point(aes(x = purple_count, y = Kelp_stipes), colour = "purple", alpha = 0.3) +
  geom_point(aes(x = red_count, y = Kelp_stipes), colour = "red", alpha = 0.3) +
    facet_wrap(vars(site_status), ncol = 1, scales = 'free') +
  ylab('Stipes (no./60m2)') + xlab('Urchins (no./60m2)')

```

urchins & kelp data distributions
```{r}
g1 <- ggplot(KelpUrchin_df) +
  stat_ecdf(aes(x = Kelp_biomasswet, colour = site_status), 
            size = 1, pad = F)

g2 <- ggplot(KelpUrchin_df) +
  stat_ecdf(aes(x = -Urchins_biomass, colour = site_status),
            size = 1, pad = F)

ggarrange(g1, g2, ncol = 1)

```

urchins & kelp v. time
```{r}
ggplot(KelpUrchin_df) +
  geom_point(aes(y = Urchins_biomass, x = year), colour = "orange", alpha = 0.3) +
  geom_point(aes(y = Kelp_biomasswet, x = year), colour = "darkgreen", alpha = 0.3) 

ggplot(KelpUrchin_df, aes(x = year)) +
  geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
  geom_smooth(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
  geom_hline(yintercept = 204, linetype = "dashed", color = "grey80", size = 1) + # ling threshold upper
  geom_hline(yintercept = 156, linetype = "dashed", color = "grey50", size = 1) + # ling threshold mean
  geom_hline(yintercept = 108, linetype = "dashed", color = "grey80", size = 1) + # ling threshold lower
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red", size = 1) + # heatwave
  geom_vline(xintercept = 2016, linetype = "dashed", color = "red", size = 1) +
  facet_wrap(vars(site_status, site)) +
  scale_y_continuous(name = "biomass (kg/60m2)", limit = c(0,1000)) +
  theme_bw()

# Anacapa East over-time averages (first by year, than over)
KelpUrchin_df %>% 
  filter(site == "ANACAPA_EAST_ISLE") %>% 
  group_by(year) %>% 
  summarise(across(Kelp_plants:Urchins_biomass, ~mean(.))) %>% 
  summarise(across(Kelp_plants:Urchins_biomass, ~mean(.)))


ggplot(KelpUrchin_df, aes(x = year)) +
  geom_point(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = Urchins_count), colour = "orange", alpha = 0.3) +
  geom_smooth(aes(y = Urchins_count), colour = "orange", alpha = 0.3) +
  facet_wrap(vars(site_status, site)) +
  scale_y_continuous(name = "Abundance (no./60m2)")


ggplot(KelpUrchin_df) +
  geom_point(aes(y = Kelp_biomasswet, x = year), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = purple_biomass, x = year), colour = "purple", alpha = 0.3) +
  geom_point(aes(y = red_biomass, x = year), colour = "red", alpha = 0.3) +  
  facet_wrap(~site)


```


Site scale
```{r}
# mean and sd across sites
KelpUrchin_df_site %>% ungroup() %>% 
  summarise(across(c(Kelp_biomasswet,Urchins_biomass), list(mean = mean, sd = sd)))

# mean and sd across forested sites only
KelpUrchin_df_site %>% ungroup() %>% 
  filter(Kelp_biomasswet>0) %>% 
  summarise(across(c(Kelp_biomasswet,Urchins_biomass), list(mean = mean, sd = sd)))

# mean and sd across sites by status
KelpUrchin_df_site %>% group_by(site_status) %>% 
  summarise(across(c(Kelp_biomasswet,Urchins_biomass), list(mean = mean, sd = sd)))


ggscatterhist(
  KelpUrchin_df_site, x = "Urchins_biomass", y = "Kelp_biomasswet",
  color = "site_status", fill = "site_status",
  size = 3, alpha = 0.6,
  margin.plot = "boxplot", ggtheme = theme_bw(), 
  xlab = 'Observed urchin biomass (kg.60m2)', ylab = 'Observed kelp biomass (kg.60m2)')


# (per m2)
 KelpUrchin_df_site <- KelpUrchin_df_site %>% 
            mutate(Urchins_biomass_m = Urchins_biomass/60, 
                   Kelp_biomasswet_m = Kelp_biomasswet/60)
ggscatterhist(
  KelpUrchin_df_site, x = "Urchins_biomass_m", y = "Kelp_biomasswet_m",
  size = 3, alpha = 0.6,
  margin.plot = "boxplot",
  ggtheme = theme_bw())



ggplot(KelpUrchin_df_site, aes(x = year)) +
  geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
  geom_smooth(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1) +
  geom_hline(yintercept = (2.6+0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold upper
  geom_hline(yintercept = 2.6*60, linetype = "dashed", color = "grey50", size = 1) + # ling threshold mean
  geom_hline(yintercept = (2.6-0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold lower  
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red", size = 1) + # heatwave
  geom_vline(xintercept = 2016, linetype = "dashed", color = "red", size = 1) +
  scale_y_continuous(name = "biomass (kg/60m2)") +
  theme_bw()



KelpUrchin_df_sitemeanMPA <- KelpUrchin_df_site %>% 
  group_by(site_status, year) %>% 
  summarise(across(Kelp_plants:Urchins_biomass, ~mean(.))) 

ggplot(KelpUrchin_df_sitemeanMPA,aes(x = year)) +
  geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", size = 2) +
  geom_point(aes(y = Urchins_biomass), colour = "orange", size = 2) +
  facet_wrap(vars(site_status), ncol = 1) +
  geom_hline(yintercept = (2.6+0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold upper
  geom_hline(yintercept = 2.6*60, linetype = "dashed", color = "grey50", size = 1) + # ling threshold mean
  geom_hline(yintercept = (2.6-0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold lower  
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red", size = 1) + # heatwave
  geom_vline(xintercept = 2016, linetype = "dashed", color = "red", size = 1) +
  scale_y_continuous(name = "biomass (kg/60m2)") +
  theme_bw()

ggplot(KelpUrchin_df_sitemeanMPA) +
  geom_point(aes(y = Kelp_biomasswet, x = Urchins_biomass, colour = site_status), size = 2) +
  geom_point(data = KelpUrchin_df_site, aes(y = Kelp_biomasswet, x = Urchins_biomass, colour = site_status), alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1) +
  geom_vline(xintercept = (2.6+0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold upper
  geom_vline(xintercept = 2.6*60, linetype = "dashed", color = "grey50", size = 1) + # ling threshold mean
  geom_vline(xintercept = (2.6-0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold lower  
  scale_y_continuous(name = "biomass (kg/60m2)") +
  theme_bw()





ggplot(KelpUrchin_df_site, aes(x = year)) +
  geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = purple_biomass), colour = "purple", alpha = 0.3) +
  geom_smooth(aes(y = purple_biomass), colour = "purple", alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1) +
  scale_y_continuous(name = "biomass (kg/60m2)") +
  theme_bw()



ggplot(KelpUrchin_df_site, aes(x = year)) +
  geom_point(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = Urchins_count), colour = "orange", alpha = 0.3) +
  geom_smooth(aes(y = Urchins_count), colour = "orange", alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1) +
  scale_y_continuous(name = "density (no./60m2)",  limits = c(0,300)) +
  theme_bw()


ggplot(KelpUrchin_df_site, aes(x = year)) +
  geom_point(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_smooth(aes(y = Kelp_stipes), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = purple_count), colour = "purple", alpha = 0.3) +
  geom_smooth(aes(y = purple_count), colour = "purple", alpha = 0.3) +
  facet_wrap(vars(site_status), ncol = 1) +
  scale_y_continuous(name = "density (no./60m2)",  limits = c(0,300)) +
  theme_bw()



```


### Fish (sheephead)
```{r}
# Fish ---

# sheephead from fish data
KFFish_SheepA <- KelpForest_Fish %>% filter(classcode == "SPUL")

# # filter small fish
# KFFish_Sheep <- KFFish_Sheep %>% filter(fish_tl <= 20)

# get weight (kg) from length-weight relationship
# SH para values from Alonzo et al (2004)
KFFish_Sheep <- KFFish_SheepA %>% 
  mutate(weight = 2.7*10^-5*fish_tl^2.86)

# remove records with na size (8/16,392)
KFFish_Sheep <- KFFish_Sheep %>% 
  filter(!is.na(weight))

# sum for each transect 
KFFish_Sheep <- KFFish_Sheep %>% select(year:fish_tl, weight) %>%
  group_by(year, site, zone, transect) %>%
  mutate(SH_biomass = count*weight) %>%
  summarise(SH_count = sum(count), SH_biomass = sum(SH_biomass))

# add site satus
KFFish_Sheep <- KFFish_Sheep %>% 
  left_join(KelpForest_Sites %>% select(site, site_status) %>% unique(),
            by = "site")

# pull out islands and separate out side
Sheep_df <- KFFish_Sheep %>% rowwise() %>% 
  filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% 
  separate_wider_regex(site,c(site = ".*", "_", side = ".*"))

# only use same sites as in urchin-data (with >2 years)
Sheep_df <- Sheep_df %>% 
  filter(site %in% KelpUrchin_df$site)

# calc stats at site level
Sheep_df_site <- Sheep_df %>% 
  group_by(year, site, site_status) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(., na.rm = T)))

# take log
Sheep_df_site <- Sheep_df_site %>% 
  mutate(across(SH_count:SH_biomass, ~log(.), .names = "{col}_ln"))
  
```


#### Sheephead recruits
- Getting value for temporal variance in recruits
- Note that sheephead don't show up in SMURF data, so we need to use recruits as a proxy for incoming larvae.
```{r}
# filter small fish
KFFish_SheepR <- KFFish_SheepA %>% filter(fish_tl <= 20)

# pull out islands and separate out side
KFFish_SheepR <- KFFish_SheepR %>% rowwise() %>% 
  filter(str_detect(site, 'ANACAPA_.*|SCI_.*|SMI_.*|SRI_.*')) %>% 
  separate_wider_regex(site,c(site = ".*", "_", side = ".*"))

# sum up over transect
KFFish_SheepR <- KFFish_SheepR %>% 
  group_by(year, site, side, zone) %>% 
  summarise(count= sum(count))

# Calc year-to-year mean/stdev across all sites
KFFish_SheepRyr <- KFFish_SheepR %>% 
  group_by(year) %>% 
  summarise(across(count, list(mean = mean, sd = sd)))

# plot
ggplot(KFFish_SheepR %>% group_by(site, year) %>% 
      summarise(across(count, list(mean = mean)))) +
  geom_line(aes(y = count_mean, x = year, color = site )) +
  theme(legend.position="none") +
  geom_line(data = KFFish_SheepRyr,
            aes(y = count_mean, x = year), size = 1) +
  theme(legend.position="none")


#calc yr2yr var ----
KFFish_SheepRyr %>% 
  ungroup %>% 
  summarise(across(count_mean:count_sd, list(mean = mean, sd = sd)))


# detrend:
    # fit linear model
    Model <- glm(KFFish_SheepRyr$count_mean ~ KFFish_SheepRyr$year)
    
    # mean plus residuals
    tsDT <- mean(KFFish_SheepRyr$count_mean) + Model$residuals
    
    # make dataframe
    Model_df <- tibble(year = KFFish_SheepRyr$year, tsDT)
    
    ggplot(Model_df) + 
      geom_line(aes(y = tsDT, x = year))

# normalise and get varaince
    tsDTnorm <- tsDT/mean(tsDT)   
    
    sd(tsDTnorm)
    
  

```


### fish, kelp, urchins 
```{r}
# v time

  # transect level
  ggplot(KelpUrchin_df, aes(x = year)) +
    geom_point(data = Sheep_df, aes(y = SH_biomass*1000, x = year), colour = "blue", alpha = 0.1) +
    geom_smooth(data = Sheep_df, aes(y = SH_biomass*1000, x = year), colour = "blue", alpha = 0.1) +
    geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
    geom_smooth(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
    geom_point(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
    geom_smooth(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
    facet_wrap(vars(site_status, site)) +
    scale_y_continuous(name = "Kelp biomass and Urchin Biomass (kg/60m2)", 
                       sec.axis= sec_axis(~./1000, name = "Sheephead biomass (kg/60m2)")) 

# Anacapa East over-time averages (first by year, than over)
Sheep_df %>% 
  filter(site == "ANACAPA_EAST_ISLE") %>% 
  group_by(year) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.))) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.)))
  
  
  # site level
  ggplot(KelpUrchin_df_site, aes(x = year)) +
    geom_point(data = Sheep_df_site, aes(y = SH_biomass*500, x = year), colour = "blue", alpha = 0.1) +
    geom_smooth(data = Sheep_df_site, aes(y = SH_biomass*500, x = year), colour = "blue", alpha = 0.1) +
    geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
    geom_smooth(aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
    geom_point(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
    geom_smooth(aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
    facet_wrap(vars(site_status), ncol = 1) +
    # facet_wrap(vars(site_status, site)) +
    scale_y_continuous(name = "Kelp biomass and Urchin Biomass*10 (kg/60m2)", limits = c(0,2000),
                       sec.axis= sec_axis(~./500, name = "Sheephead biomass (kg/60m2)")) +
    theme_bw()
  
  
  # all sites average
  Sheep_df_site %>% 
  group_by(year, site_status) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.))) %>% 
  group_by(site_status) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.)))
  
  SH_siteavg_MPA <- Sheep_df_site %>% 
  group_by(year, site_status) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.))) 
  
  ggplot(SH_siteavg_MPA) +
    geom_point(aes(y = SH_biomass, x = year, color = site_status)) +
    facet_wrap(vars(site_status), ncol = 1) +
    theme_bw()
  
  
  ggplot(KelpUrchin_df_sitemeanMPA,aes(x = year)) +
  geom_point(aes(y = Kelp_biomasswet), colour = "darkgreen", size = 3) +
    geom_point(data = KelpUrchin_df_site, aes(y = Kelp_biomasswet), colour = "darkgreen", alpha = 0.3) +
  geom_point(aes(y = Urchins_biomass), colour = "orange", size = 3) +
    geom_point(data = KelpUrchin_df_site, aes(y = Urchins_biomass), colour = "orange", alpha = 0.3) +
  geom_point(data = SH_siteavg_MPA, aes(y = SH_biomass*500, x = year),colour = "blue", size = 3) +
    geom_point(data = Sheep_df_site, aes(y = SH_biomass*500, x = year), colour = "blue", alpha = 0.1) +
  facet_wrap(vars(site_status), ncol = 1) +
  geom_hline(yintercept = (2.6+0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold upper
  geom_hline(yintercept = 2.6*60, linetype = "dashed", color = "grey50", size = 1) + # ling threshold mean
  geom_hline(yintercept = (2.6-0.8)*60, linetype = "dashed", color = "grey80", size = 1) + # ling threshold lower  
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red", size = 1) + # heatwave
  geom_vline(xintercept = 2016, linetype = "dashed", color = "red", size = 1) +
  scale_y_continuous(name = "Kelp biomass and Urchin Biomass*10 (kg/60m2)", limits = c(0,2000),
                       sec.axis= sec_axis(~./500, name = "Sheephead biomass (kg/60m2)")) +
  theme_bw()
  
  #avg last 10 yrs
  SH_siteavg_MPA %>% filter(year>2008) %>% 
  group_by(site_status) %>% 
  summarise(across(SH_count:SH_biomass, ~mean(.)))  
  

```
  
  
```{r}
# add urchins and sheephead and kelp 
# replace NAs with zero
  KUS_df_site = left_join(KelpUrchin_df_site,
                          Sheep_df_site,
            by = c("year", "site", "site_status")) %>% 
  mutate(Kelp_bio_rescaled = if_else(Kelp_biomasswet < 100, 100, Kelp_biomasswet)) %>% 
  replace(is.na(.),0)

# calc if in kelp forest or barren
# kelp < 1kg.ha = 0.006kg.60m2 = barren
  KUS_df_site <- KUS_df_site %>% 
    mutate(kelp_state = ifelse(Kelp_biomasswet >0.006, 'Forest', 'Barren'))


# save
# save(KUS_df_site, file = "PISCOdata_KUS_df_site.rData")
```

  
```{r}
  ggplot(KUS_df_site, aes(y = Urchins_biomass, x = SH_biomass, color = site_status)) +
  geom_point(alpha = 0.7) +
  facet_wrap(vars(site_status), ncol = 1, scales = 'free') + 
  geom_rug(linewidth=0.1) +
  xlab('Sheephead (kg/60m2)') + ylab('Urchins (kg/60m2)')
  
  ggplot(KUS_df_site %>% mutate(ACE = site == "ANACAPA_EAST_ISLE"), 
         aes(y = Urchins_biomass, x = SH_biomass, color = ACE)) +
     geom_point(alpha = 0.7) +
     facet_wrap(vars(site_status), ncol = 1, scales = 'free') + 
     geom_rug(linewidth=0.1) +
     xlab('Sheephead (kg/60m2)') + ylab('Urchins (kg/60m2)')
  
  ggplot(KUS_df_site, 
         aes(y = Urchins_biomass, x = SH_biomass, color = Kelp_biomasswet, shape = kelp_state)) +
    geom_point(size = 2) +
    facet_wrap(vars(site_status), ncol = 1) + 
    scale_color_viridis_c(option = "D", direction = 1) + 
    scale_shape_manual(values=c(1,16)) +
    geom_rug(linewidth=0.1) +
    xlab('Sheephead (kg/60m2)') + ylab('Urchins (kg/60m2)')+
    theme_bw()

```







